---
title: "SGLT2i-induced DKA: GWAS power / sample size calculations"
author: "Jason Moggridge"
date: "`r Sys.Date()`"
output: 
  html_document:
    highlight: haddock
---

### Information

> *"The sample size of 60 individuals with SGLT2 inhibitor DKA provides >80% power to uncover a significant association (P<5x10-8) with genetic variants (minor allele frequency 0.20; odds-ratio>5) and DKA, as calculated using the R package, genpwr. Our plan is to match cases N=60 3:1 (or so) thus we will have 180 controls."*

- **Design**: Case-control GWAS. 

- **Sample size**: 60 cases, 180+ controls. 

- **Assay**: Illumina GSA-24 (~654,027 markers)

- **Assumptions**: SGLT2-DKA associated variants are presumed to have a large effect size (*OR > 5*) and be common (*MAF 0.2*) in the study population.

- **Alpha**: the significance threshold is *0.05 / 1M tests = 5x10^-8*



### Further Assumptions*

**Genetic model**: Additive



### Code

```{r}
library(tidyverse)
library(genpwr)
```

#### Replicate the calculation of sample size: 60 cases and 180 controls...

```{r}
sample_calc <- 
  genpwr::genpwr.calc(
    calc = 'ss',
    model = 'logistic',
    ge.interaction = NULL,
    Power = .8,
    MAF = .2,
    OR = 5,
    Alpha = .05,
    Case.Rate = seq(.01, .5, .01)
  )

sample_calc |> 
    mutate(
    test = str_glue('test = {Test.Model}'),
    true = str_glue('true = {True.Model}'),
  ) |> 
  ggplot(aes(Case.Rate, N_total_at_Alpha_0.05)) +
  geom_line(alpha = .76) +
  geom_hline(yintercept = 240, lty = 2, alpha = .7) +
  geom_vline(xintercept = .25, lty = 2, alpha = .7) +
  scale_y_log10() +
  scale_color_viridis_c() + 
  facet_grid(test ~ true)
```

So we are 80% powered to detect DKA-associated markers (having MAF = 0.2 and OR > 5), using a sample of 60 cases & 180 controls, with either additive or dominant alleles, though we appear to be under-powered to detect recessive alleles. (*I am just trying to figure out how Dr. Drogmuller arrived at the sample size figure here*).

----

Say we start adding controls (with MAF=.2), how does our ability to detect variants with smaller effect sizes improve while keeping 80% power (ie. 20% type II error rate and 5% type I error rate)?

```{r}
# add controls and compute ratio for power calc
sample_sizes <-
  tibble(
    cases = 60,
    controls = 180,
    added_controls = 2 ^ seq(-1, 16, 1) |> round(),
    N_total = cases + controls + added_controls,
    k = (N_total - cases) / cases,
    Case.Rate = cases / N_total
  ) |>
  distinct()

# compute power using each (N and case rate) pair
gwas_power <- genpwr::genpwr.calc(
  calc = 'power',
  model = 'logistic',
  ge.interaction = NULL,
  N = sample_sizes$N_total,
  Case.Rate = sample_sizes$Case.Rate,
  MAF = .2,
  OR = 5,
  Alpha = 0.05,
  True.Model = c('Dominant', 'Recessive', 'Additive'),
  Test.Model = c('Dominant', 'Recessive', 'Additive', '2df')
) |>
  as_tibble() |> 
  print()
```



----

```{r}
sample_sizes |>
  left_join(
    gwas_power, by = join_by(N_total, Case.Rate)
  ) |>
  mutate(
    test = str_glue('test = {Test.Model}'),
    true = str_glue('true = {True.Model}'),
  ) |> 
  ggplot(aes(
    x = added_controls + 1,
    y = `Power_at_Alpha_0.05`,
    # color = as_factor(OR),
  )) +
  geom_point(size = .81) +
  geom_line() +
  facet_grid(test~true) +
  scale_color_viridis_d(
    end = .9, 
    option = 'mako'
  ) +
  scale_y_continuous(
    labels = scales::label_percent()
  ) +
  scale_x_log10() +
  labs(
    x = 'Additional controls (N + 1)',
    y = expression('Power at alpha = 0.05'),
    title = 
      'Power to detect associations with 60 cases and 180+ controls',
  ) +
  guides(
    shape = guide_legend('MAF', 
                         override.aes = list(size = 2, linewidth = 1))
  ) +
  theme_bw() +
  theme(
    axis.title.y = ggtext::element_markdown()
  )
```

----



```{r}
dat <- genpwr.calc(
  calc = "power", 
  model = "logistic", 
  ge.interaction = NULL,
  N = sample_sizes$N_total,
  Case.Rate = sample_sizes$Case.Rate,
  k = NULL,
  MAF = seq(0.05, 0.8, 0.05),
  OR = 5,
  Alpha = 0.05,
  True.Model = c("Dominant", "Recessive", "Additive"),
  Test.Model = c("Dominant", "Recessive", "Additive", "2df")
  ) |> 
  glimpse() |> 
  arrange(MAF) |> 
  mutate(
    test = str_glue('test = {Test.Model}'),
    true = str_glue('true = {True.Model}'),
  )

dat |> 
  ggplot(aes(MAF, Power_at_Alpha_0.05, color = MAF)) +
  geom_line(aes(group = MAF)) +
  geom_hline(aes(yintercept = .8),
             lty = 2, alpha = .2) + 
  scale_color_viridis_c(option = 'G', direction = -1) +
  facet_grid(test ~ true) +
  theme_classic() +
  labs(
    x = 'Minor Allele Frequency',
    y = 'Power at type I error rate of 0.05',
    subtitle = 
      "N Cases = 60; N Controls = 180+; OR ≥ 5",
    title = 
      "Power to detect genetic associations for varied MAF"
  )
```


<!-- New viridis palettes!! -->

<!-- ```{r} -->
<!-- c('rocket', 'mako', 'turbo') |>  -->
<!--   map( -->
<!--     ~ ggplot(diamonds, aes(carat, price, color = x)) + -->
<!--       geom_point(alpha =.5, size = .1) + -->
<!--       scale_color_viridis_c(option = .) + -->
<!--       labs(title = .) -->
<!--   ) |>  -->
<!--   patchwork::wrap_plots(nrow = 1) -->

<!-- ``` -->

what sample size is needed to detect association with an OR > 5 for MAF 0.2

```{r}
genpwr::genpwr.calc(
  calc = 'ss',
  model = 'logistic',
  Power = .8,
  MAF = .2,
  OR = 5,
  Case.Rate = 1/3
  )
```

